Left-Linear Rewriting in Adhesive Categories

### 2024-06-26 concur/fusione.tex
switches
incorporate review

\documentclass[a4paper,UKenglish,cleveref,pdftex,thm-restate,numberwithinsect,anonymous]{lipics}
\newif\ifreport
\reporttrue 
%\reportfalse

@@

\ifreport
\nolinenumbers %uncomment to disable line numbering
\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
\else
%\relatedversion{An extended version is available at \url{https://arxiv.org/abs/2007.04213}.} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}
\fi

@@

\usepackage{hyperref}
\usepackage[%disable,%
textsize=tiny]{todonotes}
%\usepackage{etex}
\usepackage[all]{xy}
\SelectTips{cm}{}
% derivations and labels
\usepackage{proof}
\newcommand{\lab}[1]{\ensuremath{\mathsf{{#1}}}}
\newcommand{\slab}[1]{\ensuremath{\scriptstyle{\mathsf{{#1}}}}}
\usepackage{wrapfig}

@@

\title{Left-linear rewriting in adhesive categories}

% fake author for anonymous submission
\author{Alan Turing} 
{Department of Mathematics, Somewhere}
{alan@turing.com}{}{}

% \author{Paolo Baldan} 
% {Department of Mathematics, University of Padova, Italy}
% {baldan@math.unipd.it}{}{}
% \author{Davide Castelnovo}
% {Department of Mathematics, University of Padova, Italy}
% {davide.castelnovo@math.unipd.it}{}{}
% \author{Andrea Corradini}
% {Department of Computer Science, University of Pisa, Italy}
% {andrea.corradini@unipi.it}{}{}

% \author{Fabio Gadducci}
% {Department of Computer Science, University of Pisa, Italy}
% {fabio.gadducci@unipi.it}{}{}

@@

\begin{abstract}
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in
  parallel. Equivalences identifying rewriting sequences which differ
  only for independent steps are at the core of the theory of
  concurrency of many formalisms.
  % 
  We investigate the issue in the context of the double pushout
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  While a consolidated theory exists for linear rules, which
  can consume, preserve and generate entities, this paper focuses on
  left-linear rules which may also ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since % the canonical notion of independence used for linear rewriting
  % systems does not work out of the box, but it requires the
  % restriction to well-behaved classes of left-linear rules.
  % %
  % Moreover,
  a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems, and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, the equivalence 
captures when 
%sequential 
steps are causally unrelated and thus could be executed in any 
order, and possibly in parallel. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}.
%
Concerning formalisms based on the rewriting paradigm,
% i.e.~where the behaviour is dictated by a set of rules that have to be instantiated with the system at hand,
concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
Noteworthy examples are
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph is used to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and 10to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered, and some new
entities can be generated. Concerning concurrency, the chosen notion
is switch (also referred to as shift) equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Chapter 3, Section~3.4]{CorradiniMREHL97}. %cite[Section~4]{Ehrig78}

@@

As an initial step, % in performing this exploration, 
we observe that for
left-linear rules some fundamental properties that play a role
in the concurrency theory of linear rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and the Church-Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways.
%
As we will observe, this prevents the development of a sensible theory
of concurrency, since switches are used as a basis for equating
sequences of rewriting steps that differ only for the order of
independent steps.

@@

\emph{Synopsis.}
% The paper has the following structure. 
In \S\ref{sec:ade} we recall the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
In \S\ref{sec:equi} we present the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define a subclass of left-linear rewriting systems
that we call well-switching rewriting systems,
and prove that all the classical results holding for linear rewriting systems
can be recovered for such a class.
Finally, in \S\ref{conclusions} we outline possible directions for future work.
%
The appendices
% collect additional data,
recall basic results on adhesive categories (Appendix~\ref{app:ade}),
present \emph{strong enforcing} rewriting systems, a class of systems
where the local Church-Rosser Theorem holds (Appendix~\ref{app:fill}),
and give full proofs of the results
(Appendix~\ref{omitted}).

@@

\begin{example}
  \label{rem:iso}
  Let $\mathsf{Iso(\X)}$ be the class of 
  isomorphisms. Any category $\X$ is $\mathsf{Iso(\X)}$-adhesive.
\end{example}

@@

\begin{theorem}
  \label{thm:slice-functors}
  Let $\X$ be an $\mathcal{M}$-adhesive category. Given an object $X$
  the category $\X/X$ is $\mathcal{M}/X$-adhesive with
  $\mathcal{M}/X:=\{m\in \mathcal{A}(\X/X) \mid m\in
  \mathcal{M}\}$. Similarly, $X/\X$ is $X/\mathcal{M}$-adhesive with
  $X/\mathcal{M}:=\{m\in \mathcal{A}(X/\X) \mid m\in \mathcal{M}\}$.

  Moreover for every small category $\Y$, the category $\X^\Y$ of
  functors $\Y\to \X$ is $\mathcal{M}^{\Y}$-adhesive, where
  $\mathcal{M}^{\Y}:=\{\eta \in \mathcal{A}(\X^\Y) \mid \eta_Y \in
    \mathcal{M} \text{ for every } Y\in \Y\}$.
\end{theorem}

@@

\begin{definition}[Derivation]
  Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X, \R)$ a
  left-linear rewriting system.  Given $G, H \in \X$, a
  \emph{derivation} $\der{D}$ with source $G$ and target $H$, written
  $\der{D}: G \Mapsto H$, is defined as a sequence
  $\{\dder{D}_i\}_{i=0}^n$ of direct derivations such that
  $\dder{D}_i : G_i \Mapsto G_{i+1}$ is a direct derivation for every
  index $i$ and $G_0=G$, $G_{n+1}=H$.
  % \begin{enumerate}
  % \item $\dder{D}_i : G_i \Mapsto G_{i+1}$ is a direct derivation for
  %   every index $i$;
  % \item $G_0=G$ and $G_{n+1}=H$.
  % \end{enumerate}
  Moreover, we have an \emph{empty derivation} $G : G \Mapsto G$ for each $G \in \X$.
  %
  We denote by $\lgh(\der{D})$ the \emph{length} of derivation $\der{D}$.
 %
  \full{We will call the number $n+1$ the \emph{length} of the derivation,
  denoted by $\lgh(\der{D})$. We also say that an empty
  derivation has length $0$.}

@@

@@

As we mentioned, for linear rewriting systems
%when viewing  derivations as concurrent computations, 
it is canonical to identify
derivations that are equal ``up to switching'', i.e.~that differs
only for the order of independent steps. The same notion can be given
for left-linear rewriting systems by relying on the notion of switch.

We first recall some %concepts and
notation regarding permutations. 
% \begin{definition}
A \emph{permutation} on
$\interval[0]{n}$ is a bijection
$\sigma : \interval[0]{n} \to \interval[0]{n}$. We call it a
\emph{transposition} when there are $i, j \in \interval{n}$,
$i \neq j$ such that $\sigma(i)=j$, $\sigma(j) = i$, and
$\sigma(k) = k$ otherwise. Such a transposition is
% normally
denoted as
$\transp{i}[j]$ and we use letter $\nu$ for transpositions.
% 
Given a permutation $\perm$ on $\interval[0]{n}$, an \emph{inversion}
for $\sigma$ is a pair $(i,j)$ such that $i<j$ and
$\sigma(j)< \sigma(i)$. We write $\inv{\sigma}$ for the set of
inversions of $\perm$.
%\end{definition}

@@

\begin{example}
  \label{ex:diff1}
  Consider the poset $(P, \sqsubseteq)$ in Fig.~\ref{fig:diff1:P} where
  $P = \mathbb{N} \cup \{a,b,c\}$ and $\sqsubseteq$ is defined by
  $m \sqsubseteq n$ if $m \geq n$, $a \sqsubseteq x$ for all $x \in P$
  and $b, c \sqsubseteq n$ for all $n \in \mathbb{N}$.
  %
  Let $\X$ be the  
  category associated with such order, which by \Cref{rem:iso} is
  $\mathsf{Iso(\X)}$-adhesive. 
  %
  Consider a rewriting
  system whose set of rules contains the following 
  \[\xymatrix{a & a \ar[r]^{\sqsubseteq} \ar@{>->}[l]_{\id{a}} & 0 & a & a
      \ar[r]^{\sqsubseteq} \ar@{>->}[l]_{\id{a}} & b }\]
  We can then consider the  derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ in Fig.~\ref{fig:diff1:rew}.
  % \[
  %   \xymatrix@C=30pt{a \ar[d]_{m_0}&& a \ar[d]_{k_0}\ar[ll]_{\id{a}}
  %     \ar[r]^{(a,0)} & 0 \ar@/^.35cm/[drrr]|(.315)\hole_(.45){\id{0}}
  %     \ar[dr]|(.3)\hole_{\id{0}} && a \ar@/_.35cm/[dlll]^(.45){(a,c)}
  %     \ar[dl]|(.3)\hole^{(a,0)}& a \ar[d]^{(a,0)}\ar@{>->}[l]_{\id{a}}
  %     \ar[rr]^{(a,b)} && b \ar[d]^{(b,0)}\\c &&
  %     \ar@{>->}[ll]^{\id{c}} c \ar[rr]_{(c,0)}&& 0 &&
  %     \ar@{>->}[ll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0}\]
  %
  Note that the two direct derivations are sequential
  independent. However there is no switch since the rule applied by
  $\dder{D}_1$ cannot be applied to $c$. In fact, there is a unique morphism
  $a\to c$, yielding the diagram in Fig.~\ref{ex:diff1:no-step}. But $b$ and $c$
  do not have a supremum in the poset underlying $\X$, thus the arrows
  $a\to b$ and $a\to c$ do not have a
  pushout. Hence we do not get a direct derivation from $c$.
\end{example}

@@

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  for $i \in \{0,1\}$ and the mediating arrows $u_i\colon K_i\to P$ 
  for $i \in \{0,1\}$ 
  %and $u_1\colon K_1\to P$ 
  into the pullback object (see \Cref{prop:tec} for details). 
  % given by \Cref{prop:tec} .
  %
  We say that $(i_0, i_1)$ is a \emph{strong independence pair} if
  the first two squares depicted  
  below are pushouts and if the pushout of $r_1 : K_1 \to R_1$ and $u_1 : K_1 \to P$ exists
  \[
    \xymatrix@R=16pt{
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}
       &K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0}
      \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0
       & P \ar@{.>}[r]_{q_1} & Q_1
    }
  \]
\end{definition}

@@

We can identify a large class of adhesive categories such that all
left-linear rewriting systems over such categories are strong
enforcing. Such class includes $\cat{Set}$ and it is closed
by comma and functor category constructions.  As such, it includes
essentially all categories (e.g., presheaves over set) that are
typically considered for modelling purposes. Notably, it contains the
category $\cat{Graph}$ of directed graphs.
This is a natural generalisation from adhesive to $\mathcal{M}$-adhesive categories of a class studied in \cite{baldan2011adhesivity} (see \Cref{app:fill} for details).

@@

\begin{example}[Non-strong enforcing left-linear rewriting system]
  \label{ex:diff2}
  In light of \Cref{pr:church}, \Cref{ex:diff1} provides an example of
  an independence pair which is not strong. This gives an example of a
  left-linear rewriting system in a $\mathcal{M}$-adhesive category
  pathological since $\mathcal{M}=\mathsf{Iso(\X)}$, but this is expected, 
  since all natural examples seem to belong to the well-behaved class mentioned above. 
 % .\todo{P: Scritto qualcosa, probabilmente da migliorare \\ D: per me va bene così}
%  pathological since $\mathcal{M}=\mathsf{Iso(\X)}$, but it needs to be so, since natural examples are all in the well-behaved class mentioned above. 
 % .\todo{P: Scritto qualcosa, probabilmente da migliorare \\ D: per me va bene così}
  % Instead every
  % left-linear system in category $\cat{Graph}$ is strong enforcing, by
  % the results of Appendix~\ref{app:fill}, summarized
  % in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

Back consistency with the theory of linear rewriting systems is ensured by the fact that all linear rewriting systems are strong enforcing (see \cref{prop:equi}).

@@

Even if we work in strong enforcing rewriting systems where sequential
independence ensures switchability, when dealing with left-linear
rules there is a further -- and possibly more serious -- issue, 
i.e.~the fact that there can be more than one independence pair between
the same derivations. This hinders the very idea of using
sequential independence as a basis of a theory of concurrency for
rewriting systems, since exchanges performed using different
independence pairs may lead to derivations that are not abstraction
equivalent, and thus one would equate computations that should
definitively be taken apart, as shown in the example below.

@@

  As a consequence, $\der{D}$ and $\der{D}'$ would be switch
  equivalent, but this is not acceptable when viewing equivalence
  classes of derivations as concurrent computations: in $\der{D}$ the
  first two steps are not sequential independent, while in $\der{D}'$
  they are, intuitively because in $\der{D}$ rule $\rho_1$ uses the
  node generated by $\rho_0$ (adding a self-loop to it), while in
  $\der{D}'$ rule $\rho_1$ uses the node that was in the initial
  graph. Also observe that the graphs $G_2$ and $G_2'$ produced after
  two steps in $\der{D}$ and $\der{D'}$ are not isomorphic. From the
  technical point of view, the property of being switch equivalent is
  not closed by prefix, and this prevents to derive a sensible
  concurrent semantics: In fact
  $\der{D} = \dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ and
  $\der{D}' = \dder{D}_0'\cdot \dder{D}_1' \cdot \dder{D}_2'$ are
  switch equivalent, while if we consider the first two steps,
  derivations $\dder{D}_0 \cdot \dder{D}_1$ and
  $\dder{D}_0' \cdot \dder{D}_1'$ are not switch equivalent.
  % if we limit to the first
  %two steps, we obtain derivations $\dder{D}_0 \cdot \dder{D}_1$ and
  %$\dder{D}_0' \cdot \dder{D}_1'$ that are not switch equivalent, while
  %$\der{D} = \dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ and
  %$\der{D}' = \dder{D}_0'\cdot \dder{D}_1' \cdot \dder{D}_2'$ would
  %be.
\end{example}

>>>>

\documentclass[a4paper,UKenglish,cleveref,pdftex,thm-restate,numberwithinsect]{lipics}
\newif\ifreport
\reporttrue 
%\reportfalse

@@

\ifreport
\nolinenumbers %uncomment to disable line numbering
\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
\else
\relatedversion{An extended version is available at \url{https://arxiv.org/abs/2007.04213}.} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}
\fi

@@

\usepackage{hyperref}
\usepackage[disable,%
textsize=tiny]{todonotes}
%\usepackage{etex}
\usepackage[all]{xy}
\SelectTips{cm}{}
% derivations and labels
\usepackage{proof}
\newcommand{\lab}[1]{\ensuremath{\mathsf{{#1}}}}
\newcommand{\slab}[1]{\ensuremath{\scriptstyle{\mathsf{{#1}}}}}
\usepackage{wrapfig}

@@

\title{Left-Linear Rewriting in Adhesive Categories}

% fake author for anonymous submission
%\author{Alan Turing} 
%{Department of Mathematics, Somewhere}
%{alan@turing.com}{}{}

\author{Paolo Baldan} 
{Department of Mathematics, University of Padova, Italy}
{baldan@math.unipd.it}{}{}

\author{Davide Castelnovo}
{Department of Mathematics, University of Padova, Italy}
{davide.castelnovo@math.unipd.it}{}{}
\author{Andrea Corradini}
{Department of Computer Science, University of Pisa, Italy}
 {andrea.corradini@unipi.it}{}{}

\author{Fabio Gadducci}
{Department of Computer Science, University of Pisa, Italy}
{fabio.gadducci@unipi.it}{}{}

@@

\begin{abstract}
  When can two sequential steps performed by a computing device be
  considered (causally) independent?  This is a relevant question for
  concurrent and distributed systems, since independence means that
  they could be executed in any order, and potentially in
  parallel. Equivalences identifying rewriting sequences which differ
  only for independent steps are at the core of the theory of
  concurrency of many formalisms.
  % 
  We investigate the issue in the context of the double pushout
  approach to rewriting in the general setting of adhesive categories.
  %, a very general setting for rewriting.  
  While a consolidated theory exists for linear rules, which
  can consume, preserve and generate entities, this paper focuses on
  left-linear rules which may also ``merge'' parts of the state.  This is an apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since % the canonical notion of independence used for linear rewriting
  % systems does not work out of the box, but it requires the
  % restriction to well-behaved classes of left-linear rules.
  % %
  % Moreover,
  a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds.
  %and this requires a substantial revision of the theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems, and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

One of the key pay-off of concurrency theory is the idea that the behaviour of 
a computational device can be modelled by abstracting its sequences of steps
%performed by the device 
via a suitable equivalence. Intuitively, the equivalence 
captures when 
%sequential 
steps are causally unrelated and thus could be executed in any 
order, and possibly in parallel. 
%
Two seminal contributions to this line of research have been
Mazurkiewicz's traces~\cite{Mazurkiewicz86} and Winskel's event structures~\cite{NPW:PNES}.
%
Concerning formalisms based on the rewriting paradigm,
% i.e.~where the behaviour is dictated by a set of rules that have to be instantiated with the system at hand,
concurrency often boils down to having a notion of 
independence between two consecutive steps.
%thus viewing a sequence of steps as a concurrent computation. 
Noteworthy examples are the
interchange law~\cite{Mes92} in term rewriting and permutation 
equivalence~\cite{JJL80} in $\lambda$-calculi.

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph is used to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered, and some new
entities can be generated. Concerning concurrency, the chosen notion
is switch (also referred to as shift) equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Chapter 3, Section~3.4]{CorradiniMREHL97}. %cite[Section~4]{Ehrig78}

@@

As an initial step, % in performing this exploration, 
we observe that for
left-linear rules some fundamental properties that play a role
in the concurrency theory of linear rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and the Church-Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways.
%
As we will observe, this prevents the development of a sensible theory
of concurrency, since switches are used as a basis for equating
sequences of rewriting steps that differ only in the order of
independent steps.

@@

\emph{Synopsis.}
% The paper has the following structure. 
In \S\ref{sec:ade} we recall the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
In \S\ref{sec:equi} we present the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define a subclass of left-linear rewriting systems
that we call well-switching rewriting systems,
and we prove that all the classical results holding for linear rewriting systems
can be recovered for such a class.
Finally, in \S\ref{conclusions} we outline possible directions for future work.
%
The appendices
% collect additional data,
recall basic results on adhesive categories (Appendix~\ref{app:ade}),
present \emph{strong enforcing} rewriting systems, a class of systems
where the local Church-Rosser Theorem holds (Appendix~\ref{app:fill}),
and give full proofs of the results
(Appendix~\ref{omitted}).

@@

\begin{example}
  \label{rem:iso}
  Every category $\X$ is $\mathsf{I(\X)}$-adhesive, where $\mathsf{I(\X)}$ is the class of 
  isomorphisms.
\end{example}

@@

\begin{theorem}
  \label{thm:slice-functors}
  Let $\X$ be an $\mathcal{M}$-adhesive category. Given an object $X$
  the category $\X/X$ is $\mathcal{M}/X$-adhesive with
  $\mathcal{M}/X:=\{m\in \mor(\X/X) \mid m\in
  \mathcal{M}\}$. Similarly, $X/\X$ is $X/\mathcal{M}$-adhesive with
  $X/\mathcal{M}:=\{m\in \mor(X/\X) \mid m\in \mathcal{M}\}$.

  Moreover for every small category $\Y$, the category $\X^\Y$ of
  functors $\Y\to \X$ is $\mathcal{M}^{\Y}$-adhesive, where
  $\mathcal{M}^{\Y}:=\{\eta \in \mor(\X^\Y) \mid \eta_Y \in
    \mathcal{M} \text{ for every } Y\in \Y\}$.
\end{theorem}

@@

\begin{definition}[Derivation]
  Let $\X$ be an $\mathcal{M}$-adhesive category and $(\X, \R)$ a
  left-linear rewriting system.  Given $G, H \in \X$, a
  \emph{derivation} $\der{D}$ with source $G$ and target $H$, written
  $\der{D}: G \Mapsto H$, is defined as a sequence
  $\{\dder{D}_i\}_{i=0}^n$ of direct derivations such that
  $\dder{D}_i : G_i \Mapsto G_{i+1}$ is a direct derivation for every
  index $i$ and $G_0=G$, $G_{n+1}=H$.
  % \begin{enumerate}
  % \item $\dder{D}_i : G_i \Mapsto G_{i+1}$ is a direct derivation for
  %   every index $i$;
  % \item $G_0=G$ and $G_{n+1}=H$.
  % \end{enumerate}
  Moreover, we have an \emph{empty derivation} $G : G \Mapsto G$ for each $G \in \X$.
  %
  We denote by $\lgh(\der{D})$ the \emph{length} of a derivation $\der{D}$.
 %
  \full{We will call the number $n+1$ the \emph{length} of the derivation,
  denoted by $\lgh(\der{D})$. We also say that an empty
  derivation has length $0$.}

@@

\begin{remark}\label{rem:uni}
	It is worth mentioning that if $(\X, \R)$ is a linear rewriting system then two derivations $\dder{D}_0$ and $\dder{D}_1$ can have at most one independence pair. Indeed if $(i_0,i_1)$ and $(i'_0, i'_1)$ are independence pairs, then
	\[g_0\circ i_1 = g_0\circ i'_1 \qquad f_1\circ i_0=f_1\circ i'_0\]
But in linear systems $g_0$ and $f_1$ are both monos, entailing $i_0=i'_0$ and $i_1=i'_1$.
\end{remark}

@@

As we mentioned, for linear rewriting systems
%when viewing  derivations as concurrent computations, 
it is canonical to identify
derivations that are equal ``up to switching'', i.e.~that differ
only in the order of independent steps. The same notion can be given
for left-linear rewriting systems by relying on the notion of switch.

We first recall some %concepts and
notation regarding permutations. 
% \begin{definition}
A \emph{permutation} on
$\interval[0]{n}$ is a bijection
$\sigma : \interval[0]{n} \to \interval[0]{n}$. We call it a
\emph{transposition} when there are $i, j \in \interval{n}$,
$i \neq j$ such that $\sigma(i)=j$, $\sigma(j) = i$, and
$\sigma(k) = k$ otherwise. Such a transposition is
% normally
denoted as
$\transp{i}[j]$ and we use the letter $\nu$ for transpositions.
% 
Given a permutation $\perm$ on $\interval[0]{n}$, an \emph{inversion}
for $\sigma$ is a pair $(i,j)$ such that $i<j$ and
$\sigma(j)< \sigma(i)$. We write $\inv{\sigma}$ for the set of
inversions of $\perm$.
%\end{definition}

@@

\begin{example}
  \label{ex:diff1}
  Consider the poset $(P, \sqsubseteq)$ in Fig.~\ref{fig:diff1:P} where
  $P = \mathbb{N} \cup \{a,b,c\}$ and $\sqsubseteq$ is defined by
  $m \sqsubseteq n$ if $m \geq n$, $a \sqsubseteq x$ for all $x \in P$
  and $b, c \sqsubseteq n$ for all $n \in \mathbb{N}$.
  %
  Let $\X$ be the  
  category associated with this order, which by \Cref{rem:iso} is
  $\mathsf{I(\X)}$-adhesive. 
  %
  Consider a rewriting
  system whose set of rules contains the following 
  \[\xymatrix{a & a \ar[r]^{\sqsubseteq} \ar@{>->}[l]_{\id{a}} & 0 & a & a
      \ar[r]^{\sqsubseteq} \ar@{>->}[l]_{\id{a}} & b }\]
  We can then consider the  derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ in Fig.~\ref{fig:diff1:rew}.
  % \[
  %   \xymatrix@C=30pt{a \ar[d]_{m_0}&& a \ar[d]_{k_0}\ar[ll]_{\id{a}}
  %     \ar[r]^{(a,0)} & 0 \ar@/^.35cm/[drrr]|(.315)\hole_(.45){\id{0}}
  %     \ar[dr]|(.3)\hole_{\id{0}} && a \ar@/_.35cm/[dlll]^(.45){(a,c)}
  %     \ar[dl]|(.3)\hole^{(a,0)}& a \ar[d]^{(a,0)}\ar@{>->}[l]_{\id{a}}
  %     \ar[rr]^{(a,b)} && b \ar[d]^{(b,0)}\\c &&
  %     \ar@{>->}[ll]^{\id{c}} c \ar[rr]_{(c,0)}&& 0 &&
  %     \ar@{>->}[ll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0}\]
  %
  Note that the two direct derivations are sequential
  independent. However there is no switch since the rule applied by
  $\dder{D}_1$ cannot be applied to $c$. In fact, there is a unique morphism
  $a\to c$, yielding the diagram in Fig.~\ref{ex:diff1:no-step}. But $b$ and $c$
  do not have a supremum in the poset underlying $\X$, thus the arrows
  $a\to b$ and $a\to c$ do not have a
  pushout. Hence we do not get a direct derivation from $c$.
\end{example}

@@

  Consider the pullback of $g_0$ and $f_1$, which yields $p_i : P \to D_i$
  for $i \in \{0,1\}$ and the mediating arrows $u_i\colon K_i\to P$ 
  for $i \in \{0,1\}$ 
  %and $u_1\colon K_1\to P$ 
  into the pullback object (see \Cref{prop:tec} for details). 
  % given by \Cref{prop:tec} .
  %
  We say that $(i_0, i_1)$ is a \emph{strong independence pair} if
  the first two squares depicted  
  below are pushouts and if the pushout of $r_1 : K_1 \to R_1$ and $u_1 : K_1 \to P$ exists
  \[
    \xymatrix@R=16pt{
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}
       &K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0}
      \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0
       & P \ar@{.>}[r]_{q_1} & Q_1
    }
  \]
\end{definition}

@@

We can identify a large class of adhesive categories such that all
left-linear rewriting systems over such categories are strong
enforcing. This class includes $\cat{Set}$ and it is closed
under comma and functor category constructions.  As such, it includes
essentially all categories (e.g., presheaves over set) that are
typically considered for modelling purposes. Notably, it contains the
category $\cat{Graph}$ of directed graphs.
This is a natural generalisation from adhesive to $\mathcal{M}$-adhesive categories of a class studied in \cite{baldan2011adhesivity} (see \Cref{app:fill} for details).

@@

\begin{example}[Non-strong enforcing left-linear rewriting system]
  \label{ex:diff2}
  In light of \Cref{pr:church}, \Cref{ex:diff1} provides an example of
  an independence pair which is not strong. This gives an example of a
  left-linear rewriting system in a $\mathcal{M}$-adhesive category which is quite
  pathological since $\mathcal{M}=\mathsf{I(\X)}$. However, this is expected, 
  since all natural examples seem to belong to the well-behaved class mentioned above. 
 % .\todo{P: Scritto qualcosa, probabilmente da migliorare \\ D: per me va bene così}
%  pathological since $\mathcal{M}=\mathsf{I(\X)}$, but it needs to be so, since natural examples are all in the well-behaved class mentioned above. 
 % .\todo{P: Scritto qualcosa, probabilmente da migliorare \\ D: per me va bene così}
  % Instead every
  % left-linear system in category $\cat{Graph}$ is strong enforcing, by
  % the results of Appendix~\ref{app:fill}, summarized
  % in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

Consistency with the theory of linear rewriting systems is ensured by the fact that all linear rewriting systems are strong enforcing (see \cref{prop:equi}).

@@

Even if we work in strong enforcing rewriting systems where sequential
independence ensures switchability, when dealing with left-linear
rules there is a further -- and possibly more serious -- issue, 
namely the fact that there can be more than one independence pair between
the same derivations (cfr. \Cref{rem:uni}). This hinders the very idea of using
sequential independence as a basis of a theory of concurrency for
rewriting systems, since exchanges performed using different
independence pairs may lead to derivations that are not abstraction
equivalent, and thus one would equate computations that should
definitively be taken apart, as shown in the example below.

@@

  As a consequence, $\der{D}$ and $\der{D}'$ would be switch
  equivalent, but this is not acceptable when viewing equivalence
  classes of derivations as concurrent computations: in $\der{D}$ the
  first two steps are not sequential independent, while in $\der{D}'$
  they are, intuitively because in $\der{D}$ rule $\rho_1$ uses the
  node generated by $\rho_0$ (adding a self-loop to it), while in
  $\der{D}'$ rule $\rho_1$ uses the node that was in the initial
  graph. Also observe that the graphs $G_2$ and $G_2'$ produced after
  two steps in $\der{D}$ and $\der{D'}$ are not isomorphic. From the
  technical point of view, the property of being switch equivalent is
  not closed by prefix, and this prevents deriving a sensible
  concurrent semantics: In fact
  $\der{D} = \dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ and
  $\der{D}' = \dder{D}_0'\cdot \dder{D}_1' \cdot \dder{D}_2'$ are
  switch equivalent, while if we consider the first two steps,
  derivations $\dder{D}_0 \cdot \dder{D}_1$ and
  $\dder{D}_0' \cdot \dder{D}_1'$ are not switch equivalent.
  % if we limit to the first
  %two steps, we obtain derivations $\dder{D}_0 \cdot \dder{D}_1$ and
  %$\dder{D}_0' \cdot \dder{D}_1'$ that are not switch equivalent, while
  %$\der{D} = \dder{D}_0\cdot \dder{D}_1 \cdot \dder{D}_2$ and
  %$\der{D}' = \dder{D}_0'\cdot \dder{D}_1' \cdot \dder{D}_2'$ would
  %be.
\end{example}
